Big Ideas in Sports Analytics and Statistical Tools for their Investigation

### 2022-07-14 sports_analytics.Rmd
editing and references

---
title: "Big Ideas in Sports Analytics and Statistical Tools for their Investigation"
authors:
  - name: Benjamin S. Baumer
#    thanks: Use footnote for providing further information about author (webpage, alternative address)---*not* for acknowledging funding agencies. Optional.
    department: Statistical & Data Sciences
    affiliation: Smith College
    location: Northampton, MA 01063
    email: bbaumer@smith.edu
    ORCID: 0000-0002-3279-0516
  - name: Gregory J. Matthews
    department: Mathematics and Statistics
    affiliation: Loyola University Chicago
    location: Chicago, IL 60660
    email: gmatthews1@luc.edu
    ORCID: 0000-0002-8413-5097
  - name: Quang Nguyen
    department: Statistics & Data Science
    affiliation: Carnegie Mellon University
    location: Pittsburgh, PA 15213
    email: nmquang@cmu.edu
  - name: Katherine Evans
    affiliation: Monumental Basketball
    location: Washington, DC 20003
    email: kevans@monumentalsports.com
abstract: |
  Sports analytics---broadly defined as the pursuit of improvement in athletic performance through the analysis of data---has expanded its footprint both in the professional sports industry and in academia over the past 30 years. We connect four(?) big ideas that are common across multiple sports, and explore both the shared similarities and individual idiosyncracies of analytical approaches in each sport. While our focus is on the concepts underlying each type of analysis, any implementation necessarily involves both statistical methodologies and sources of data. Where appropriate, we outline how data, models, and knowledge of the sport combine to generate actionable insights. This paper should serve as a useful overview for anyone becoming interested in the study of sports analytics. 
keywords:
  - sports analytics
  - R packages
  - sports data
  - pairwise comparisons
  - datasets
bibliography: [refs.bib, pkgs.bib]
biblio-style: unsrt   
# use apa for WIRE
csl: apa.csl
output: rticles::arxiv_article
header-includes:
  - \usepackage{amsmath}
---

@@

For clarity, we start with baseball due to the discrete nature of the sport. 
In baseball, $s$ is typically determined by two factors: the configuration of the runners on base (there are 8 possibilities) and the number of outs (3 possibilities). 
Thus, there $|S| = 24 = 8 \cdot 3$ basic states of an inning in baseball (25, if you include the absorbing state of 3 outs), and we are often interested in the number of runs that will be scored from some state until the end of the inning. 
In this example using baseball, $\mathbb{E}[X | s] \,$ is the expected number of runs scored between now and the end of the inning given that the inning is currently in state $s$. 
The collection of estimates $\mathbb{E}[X | s]$ for all 24 states is called the *expected run matrix* [^matrix], and it is foundational in baseball analytics. 
Figure \ref{fig:lindsey} shows a reproduction of Lindsey's original calculations and Table \ref{tab:erm} shows the expected run matrix in its more familiar form.

@@

Second, the characterization of $S$ as having 24 states is only the simplest possible. 
The inning, or the score of the game, or even the weather, might reasonably be incorporated into $S$, as those conditions might reasonably affect the estimate of $\mathbb{E}[X | s]$. 
More definitively, the identity of the current batter, pitcher, or batter on deck, might also affect the estimate of $\mathbb{E}[X | s]$. 
Indeed, @Tango2007book show that when a particularly weak-hitting batter is up (i.e., the pitcher), a sacrifice bunt becomes much more effective strategy.

@@

## In other sports with discrete states

@@

@goldner2012markov and @goldner2017situational proposes a Markov model for estimating expected points in football. In particular, the author considers a football drive as an *absorbing Markov chain*, consisting of distinct *absorbing states* touchdowns, field goals, and other play outcomes. For any given play, the expected points is calculated using the absorption probabilities for different scoring events.

@@

@@

## Sports with continuous states

Even in sports where the concept of a state is not so easy to define, the value of a possession can be estimated, with the help of tracking data. 
Over the past decade or so, professional sports leagues have collected tracking data which record the locations of all players and the ball (or puck) throughout a game. 
This sophisticated type of data allows researchers to produce advanced analyses of the captured spatio-temporal information and better understand the game, which is much more difficult to be accomplished with pre-existing resources such as traditional box-score results and play-by-play data.

In basketball, @cervone2014pointwise and @cervone2016multiresolution introduce expected possession value (EPV) as an assessment of a player's on-court performance. 
This metric is a continuous-time estimate of the expected number of points for the offensive team on a given possession using player and ball locations. 
The EPV takes into account all possible outcomes (a shot attempt, a pass, etc.) for a given player with the ball, with different weights being assigned to each decision. 
Consequently, the authors derive a metric called EPV-Added (EPVA), measuring a player's EPV contribution in a given situation relative to a league-average player.

(include code demo? https://github.com/dcervone/EPVDemo)

Another framework for estimating expected points in basketball was proposed by @Sicilia2019DeepHoops. 
The authors offer a different point of view on expected points, where they first consider a classification model which returns the probabilities for whether a player would commit a foul (shooting and non-shooting), turnover, or attempt a shot. 
The values associated with each of those "terminal actions" are then utilized to compute the expected points within a basketball play.

@@

In American football, @Yurko2020going use tracking data provided by the 2019 NFL Big Data Bowl (citation) to model the expected yards gained for a ball-carrier during the course of a play. 
As an extension to pre-existing approaches, the authors use conditional density estimation to obtain a probability distribution for the number of yards gained during the play, rather than only producing a single estimate for the expected yards gained. 
Accordingly, the probability of various types of outcome at the end of a play such as a touchdown or a first-down gain can be computed from the yards gained distribution.

The notion of EPV has also been extended to soccer recently.
@fernandez2021framework implement deep learning methods to examine the instantaneous expected value of soccer possessions. 
This approach considers passes, ball drives, and shots in soccer as the main set of actions used to compute expected possession value.

@@

For example, in the bottom of the ninth inning of a tied baseball game, the optimal strategy for winning the game is maximizing the probability of scoring *at least one run*, which may differ from the strategy of maximizing expected runs. 
If we let $U$ be the set of all strategies, then we assert that it is not always the case that the strategy that maximizes the expected number of points will maximize the probability of winning:
$$
  \underset{u \in U}{\arg \max \,}{\Pr[X > 0 | s]} \neq \underset{u \in U}{\arg \max \,}{\mathbb{E}[X | s]} \,.
$$
Consider the situation where runners are on first and third base, and the score is tied in the bottom of the ninth inning with no one out. 
Table \ref{tab:tied} reveals that the expected number of runs scored in the remainder of the inning is 1.94 runs, while the the probability of scoring zero runs is 0.13. 
The defense is in a tight spot, facing an 87% probability of losing the game. 
However, by walking the hitter to load the bases, they create the opportunity to force the lead runner at home and thus reduce the chances of scoring to 82%, even though they raise the expected number of runs scored to 2.22. 
In this case, the defensive team is wise to pursue the strategy of maximizing the expected number of runs scored, because it *minimizes* the probability of scoring at least one run.

@@

A related, but different, concept is the notion of *win probability*.
Win probability is simply an estimate of the probability that a team will win the game, given its current state $s$. 
Extending the mathematical framework we defined in Section \ref{sec:ev}, let $W_i$ be a binary random variable that indicates a win for team $i$. 
Then, 
$$
  \Pr[W_i | s] \,,
$$
is the win probability for team $i$ in the state $s$.

@@

In this section, we discuss notable previous work on win probability in baseball, football, basketball, and several other sports.

@@

## Football

@@

Figure \ref{fig:nflfastr_wp} shows a win probability graph for the January 14, 2018 NFL Playoffs Divisional Round matchup between the New Orleans Saints and the Minnesota Vikings. 
We obtain the estimated probability of winning for each team using the \pkg{nflfastR} R package, which implements a gradient boosting model via the \pkg{xgboost} library (@R-xgboost) for estimating win probabilities. 
Minnesota was leading throughout the first three quarters of the game, having win probabilities of 0.869, 0.941, 0.742 at the end of the first, second, and third quarters, respectively. 
The win probabilities get close to parity late in the fourth quarter, when the Saints took the lead with 3:01 left in the game. 
The last play of this game, famously known as the [Minneapolis Miracle](https://en.wikipedia.org/wiki/Minneapolis_Miracle), resulted in a drastic swing in win probabilities for both teams. 
With 10 seconds remaining in the game, the Vikings begin the final possession with a 25.3% chance of winning. 
Their probability increased to a perfect 1 when Stefon Diggs scored a game-winning 61-yard receiving touchdown as the game clock expired.

@@

@Stern1994 uses Brownian motion to investigate the scoring process of basketball. 
Let $p(l, t)$ represent the win probability for the home team given an $l$-point lead after $t$ seconds of game time. 
The model introduced by @Stern1994 is a probit regression model, which provides an estimate for $p(l, t)$. 
In particular, 
$$
p(l, t) = \Phi\left(\frac{l + (1-t)\mu}{\sqrt{(1-t)\sigma^2}}\right)
,
$$
Here, a Brownian motion process with drift $\mu$ points advantage for the home team and variance $\sigma^2$ is used to model the score difference between the home and away teams.

@@

The idea of win probability is also applied to other sports, with a diverse range of statistical techniques being used to estimate the probability of winning for a player or team. 
@Brenzel2019 uses three-dimensional Markov models to estimate win probability throughout a curling match. 
In particular, the authors propose both a homogeneous and heterogeneous Markov models for estimating the chance of winning in curling, with different independence assumptions on the relationship between performance and the current state of the game. 
In esports, @Maymin2020 uses logistic regression to build a well-calibrated in-game win probability model for each specific moment during a game of League of Legends. 
More recently, @Guan2022 develops an in-game win probability model for the National Rugby League using functional data analysis. 
In this approach, the rugby event data are treated as functional, and the win probability is expressed as a function of the match time.

@@

Another important idea in sports analytics is the estimation of team strength. 
A popular method for estimating team strength in sports is through pairwise evaluations. 
Perhaps the most famous probability model for predicting the outcome of a paired comparison is the Bradley-Terry model (BTM) [@Bradley1952]. 
For a pair of players (or teams) $i$ and $j$, let $\Pi_{ij}$ denote the probability that $i$ is preferred to $j$. 
Then the BTM is a logistic regression model with parameters $\beta_i, \beta_j$ such that

$$
\log\left(\frac{\Pi_{ij}}{\Pi_{ji}}\right) = \beta_i - \beta_j
$$

@@

The BTM can be implemented in `R` [@Rcitation2022] via the \pkg{BradleyTerry2} package [@R-BradleyTerry2]. 
As an example, we consider the data given in @agresti2018introduction (page 247) on tennis results from 2014--2018 for five men's professional players Novak Djokovic, Roger Federer, Andy Murray, Rafael Nadal, and Stan Wawrinka. 
We fit a BTM to estimate the win probability for each pair of players and obtain a ranking for this group of five.

@@

List contributions from individuals who do not meet the criteria for authorship (for example, to recognize people who provided technical help, collation of data, writing assistance, acquisition of funding, or a department chairperson who provided general support), with permission from the individual. Thanks to anonymous reviewers are not appropriate.

>>>>

---
title: "Big Ideas in Sports Analytics and Statistical Tools for their Investigation"
authors:
  - name: Benjamin S. Baumer
#    thanks: Use footnote for providing further information about author (webpage, alternative address)---*not* for acknowledging funding agencies. Optional.
    department: Statistical & Data Sciences
    affiliation: Smith College
    location: Northampton, MA 01063
    email: bbaumer@smith.edu
    ORCID: 0000-0002-3279-0516
  - name: Gregory J. Matthews
    department: Mathematics and Statistics
    affiliation: Loyola University Chicago
    location: Chicago, IL 60660
    email: gmatthews1@luc.edu
    ORCID: 0000-0002-8413-5097
  - name: Quang Nguyen
    department: Statistics & Data Science
    affiliation: Carnegie Mellon University
    location: Pittsburgh, PA 15213
    email: nmquang@cmu.edu
abstract: |
  Sports analytics---broadly defined as the pursuit of improvement in athletic performance through the analysis of data---has expanded its footprint both in the professional sports industry and in academia over the past 30 years. We connect four(?) big ideas that are common across multiple sports, and explore both the shared similarities and individual idiosyncracies of analytical approaches in each sport. While our focus is on the concepts underlying each type of analysis, any implementation necessarily involves both statistical methodologies and sources of data. Where appropriate, we outline how data, models, and knowledge of the sport combine to generate actionable insights. This paper should serve as a useful overview for anyone becoming interested in the study of sports analytics. 
keywords:
  - sports analytics
  - R packages
  - sports data
  - pairwise comparisons
  - datasets
bibliography: [refs.bib, pkgs.bib]
biblio-style: unsrt   
# use apa for WIRE
csl: apa.csl
output: rticles::arxiv_article
header-includes:
  - \usepackage{amsmath}
---

@@

For clarity, we start with baseball due to the discrete nature of the sport. 
In baseball, $s$ is typically determined by two factors: the configuration of the runners on base (there are 8 possibilities) and the number of outs (3 possibilities). 
Thus, there $|S| = 24 = 8 \cdot 3$ basic states of an inning in baseball[^state], and we are often interested in the number of runs that will be scored from some state until the end of the inning. 
In this example using baseball, $\mathbb{E}[X | s] \,$ is the expected number of runs scored between now and the end of the inning given that the inning is currently in state $s$. 
The collection of estimates $\mathbb{E}[X | s]$ for all 24 states is called the *expected run matrix* [^matrix], and it is foundational in baseball analytics. 
Figure \ref{fig:lindsey} shows a reproduction of George Lindsey's original calculations and Table \ref{tab:erm} shows the expected run matrix in its more familiar form.

[^state]: 25, if you include the absorbing state of 3 outs.

@@

Second, the characterization of $S$ as having 24 states is only the simplest possible. 
The inning, or the score of the game, or even the weather, might reasonably be incorporated into $S$, as those conditions might reasonably affect the estimate of $\mathbb{E}[X | s]$. 
More definitively, the identity of the current batter, pitcher, or batter on deck, might also affect the estimate of $\mathbb{E}[X | s]$. 
Indeed, @Tango2007book show that when a particularly weak-hitting batter is up (i.e., the pitcher), a sacrifice bunt becomes a more effective strategy.

@@

## In American football

@@

@goldner2012markov and @goldner2017situational propose a Markov model for estimating expected points in football. 
In particular, the author consider a football drive as an *absorbing Markov chain*, consisting of distinct *absorbing states* touchdowns, field goals, and other play outcomes. 
For any given play, the expected points is calculated using the absorption probabilities for different scoring events.

@@

### 4th down strategy

@@

## In sports with continuous states

Even in sports where the concept of a state is not so easy to define, the value of a possession can be estimated with the help of tracking data. 
Over the past decade or so, professional sports leagues have collected tracking data which record the locations of all players and the ball (or puck) throughout a game. 
This high-resolution data allows researchers to produce advanced analyses of the captured spatio-temporal information and better understand the game. 
This is a great leap forward from older resources such as traditional box-score results and play-by-play data.

In basketball, @cervone2014pointwise and @cervone2016multiresolution introduce expected possession value (EPV) as means toward an assessment of a player's on-court performance. 
This metric is a continuous-time estimate of the expected number of points for the offensive team on a given possession using player and ball locations. 
The EPV takes into account all possible outcomes (a shot attempt, a pass, etc.) for a given player with the ball, with different weights being assigned to each decision. 
Consequently, the authors derive a metric called EPV-Added (EPVA), measuring a player's EPV contribution in a given situation relative to a league-average player.

(include code demo? https://github.com/dcervone/EPVDemo)
> It might be cool to include one of their graphics here, if can. Or regenerate one.

Another framework for estimating expected points in basketball is proposed by @Sicilia2019DeepHoops. 
The authors offer a different point of view on expected points, where they first consider a classification model which returns the probabilities for whether a player would commit a foul (shooting and non-shooting), turnover, or attempt a shot. 
The values associated with each of those "terminal actions" are then used to compute the expected points within a basketball play.

@@

In American football, @Yurko2020going use tracking data provided by the [2019 NFL Big Data Bowl](https://operations.nfl.com/gameday/analytics/big-data-bowl/past-big-data-bowl-recaps/) to model the expected yards gained for a ball-carrier during the course of a play. 
As an extension to pre-existing approaches, the authors use conditional density estimation to obtain a probability distribution for the number of yards gained during the play, rather than only producing a single estimate for the expected yards gained. 
Accordingly, the probability of various types of outcomes at the end of a play such as a touchdown or a first-down gain can be computed from the yards gained distribution.

The notion of EPV has also been extended to soccer.
@fernandez2021framework implement deep learning methods to examine the instantaneous expected value of soccer possessions. 
This approach considers passes, ball drives, and shots in soccer as the main set of actions used to compute expected possession value.

@@

For example, in the bottom of the ninth inning of a tied baseball game, the optimal strategy for winning the game is maximizing the probability of scoring *at least one run*, which may differ from the strategy of maximizing expected runs. 
If we let $U$ be the set of all strategies, then we assert that it is not always the case that the strategy $u$ that maximizes the expected number of points will maximize the probability of winning:
$$
  \underset{u \in U}{\arg \max \,}{\Pr[X > 0 | s, u]} \neq \underset{u \in U}{\arg \max \,}{\mathbb{E}[X | s, u]} \,.
$$
Consider the situation where runners are on first and third base, and the score is tied in the bottom of the ninth inning with no one out. 
Table \ref{tab:tied} reveals that the expected number of runs scored in the remainder of the inning is 1.94 runs, while the the probability of scoring zero runs is 0.13. 
The defense is in a tight spot, facing an 87% probability of losing the game. 
However, by walking the hitter to load the bases, they create the opportunity to force the lead runner at home and thus reduce the chances of scoring to 82%, even though they raise the expected number of runs scored to 2.22. 
In this case, the defensive team is wise to pursue the strategy of maximizing the expected number of runs scored, because it *minimizes* the probability of scoring at least one run.

@@

A related, but different concept is the notion of *win probability*.
Win probability is simply an estimate of the probability that a team will win the game, given its current state $s$. 
Extending the mathematical framework we defined in Section \ref{sec:ev}, let $W_i$ be a binary random variable that indicates a win for team $i$. 
Then, 
$$
  \Pr[W_i | s] \,,
$$
is the win probability for team $i$ in the state $s$.

@@

In this section, we discuss notable previous work on win probability in baseball, American football, basketball, and several other sports.

@@

## American football

@@

Figure \ref{fig:nflfastr_wp} shows a win probability graph for the January 14, 2018 NFL Playoffs Divisional Round matchup between the New Orleans Saints and the Minnesota Vikings. 
We obtain the estimated probability of winning for each team using the \pkg{nflfastR} R package, which implements a gradient boosting model via the \pkg{xgboost} library (@R-xgboost) for estimating win probabilities. 
Minnesota was leading throughout the first three quarters of the game, having win probabilities of 0.869, 0.941, 0.742 at the end of the first, second, and third quarters, respectively. 
The win probabilities get close to parity late in the fourth quarter, when the Saints took the lead with 3:01 left in the game. 
The last play of this game---famously known as the [Minneapolis Miracle](https://en.wikipedia.org/wiki/Minneapolis_Miracle)---resulted in a drastic swing in win probabilities for both teams. 
With 10 seconds remaining in the game, the Vikings begin the final possession with a 25.3% chance of winning. 
Their probability increased to a perfect 1 when Stefon Diggs scored a game-winning 61-yard receiving touchdown as the game clock expired.

@@

@Stern1994 uses Brownian motion to investigate the scoring process of basketball. 
Let $p(l, t)$ represent the win probability for the home team given an $l$-point lead after $t$ seconds of game time. 
The model introduced by @Stern1994 is a probit regression model, which provides an estimate for $p(l, t)$. 
In particular, 
$$
p(l, t) = \Phi\left(\frac{l + (1-t)\mu}{\sqrt{(1-t)\sigma^2}}\right)
\,.
$$
Here, a Brownian motion process with drift $\mu$ points advantage for the home team and variance $\sigma^2$ is used to model the score difference between the home and away teams.

@@

The idea of win probability is also applied to other sports, with a diverse range of statistical techniques being used to estimate the probability of winning for a player or team. 
@Brenzel2019 uses three-dimensional Markov models to estimate win probability throughout a curling match. 
In particular, the authors propose both homogeneous and heterogeneous Markov models for estimating the chance of winning in curling, with different independence assumptions on the relationship between performance and the current state of the game. 
In esports, @maymin2021smart uses logistic regression to build a well-calibrated in-game win probability model for each specific moment during a game of League of Legends. 
@Guan2022 develops an in-game win probability model for the National Rugby League using functional data analysis. 
In this approach, the rugby event data are treated as functional, and the win probability is expressed as a function of the match time.

@@

A third crucial idea in sports analytics is the estimation of team strength. 
A popular method for estimating team strength in sports is through pairwise evaluations. 
Perhaps the most famous probability model for predicting the outcome of a paired comparison is the Bradley-Terry model (BTM) [@Bradley1952]. 
For a pair of players (or teams) $i$ and $j$, let $\Pi_{ij}$ denote the probability that $i$ is preferred to $j$. 
Then the BTM is a logistic regression model with parameters $\beta_i, \beta_j$ such that

$$
\log\left(\frac{\Pi_{ij}}{\Pi_{ji}}\right) = \beta_i - \beta_j \,.
$$

@@

The BTM can be implemented in `R` [@R-base] via the \pkg{BradleyTerry2} package [@R-BradleyTerry2]. 
As an example, we consider the data given in @agresti2018introduction (page 247) on tennis results from 2014--2018 for five men's professional players: Novak Djokovic, Roger Federer, Andy Murray, Rafael Nadal, and Stan Wawrinka. 
We fit a BTM to estimate the win probability for each pair of players and obtain a ranking for this group of five.

@@

We are grateful to Michael Lopez and Katherine Evans for their thoughts on early versions of this paper.
